Evolution of ORM Systems: Formal Foundations

### 2013-01-02 doc/description.tex
changes in organization of Transformation chapter
reorganization of transformations

$\rho_p(p, d) = \begin{cases}
  \inference{cardinality(p) = 1 \wedge !\;mandatory(p)
 \\ \wedge \exists t : t = table(name(owningClass(p)), d) 
 \\ \wedge \not instanciated(owningClass(p))
 }{\begin{gathered}
	  columns(t) \cup column(name(p), \rho_t(type(p)), \varnothing )
\end{gathered}
 } \\ \\

  \inference{cardinality(p) = 1 \wedge mandatory(p)
 \\ \wedge \exists t : t = table(name(owningClass(p)), d)}{\begin{gathered}
 	 columns(t) \cup column(name(p), \rho_t(type(p)), NOTNULL ) \\ 
\end{gathered}} 
\\ \\

@@

 \inference{ cardinality(p) > 1 \wedge mandatory(p) \\ \wedge \exists t : t = table(name(owningClass(p)), d) }{\begin{gathered} tables(d) \cup  table(name(p), primaryKey("Id"), \\ column("value", \rho_t(type(p)), NOTNULL), \\ foreignKey(name(p), name(owningClass(p)), NOTNULL)
	 \end{gathered}}
 \end{cases}$

@@

\subsection{Software Creation}
There are defined three transformations which are able to create a software: $newClass$, $addProperty$, $addAssociation$. These three transformations can be used to create a software from a scratch. The first of them has no impact on stored data, whereas the rest of them can in some cases.
\begin{gather*}
\mathbf{addProperty} : Label \times Property \times State \rightarrow State \\
\Psi(s, addProperty(l, p)) = application(s) \circ addProp(l,p) \\
\Phi(s, addProperty(l, p)) = database(s) \bullet insertProp(l, p) \\
s \in Softwares, l \in Label, p \in Properties
\end{gather*}
Adding a property into an application structure has to verify a name collision only:
\begin{gather*}
addProp : Label \times Property \times Application \rightarrow Application \\
addProp(l, p, a) = \inference{\exists c \in classes(a) : name(c) = l \\ \wedge \forall q \in  properties(p) : name(q) \neq name(p)}{properties(c) \cup p } \\
l \in Label, p \in Properties, a \in Applications
\end{gather*}
Whereas the $insertProp$ on database level can change stored data too, because a mandatory property can be added into a class with already existing instances, then a default value has to be added to all existing instances.
\begin{gather*}
insertProp(p, d) = \begin{cases}
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge (!\;mandatory(p) \vee !\;instantiated(owningClass(p))}{ \rho(p,d)} 
\\ \\ 
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge mandatory(p) \wedge instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered}
 \rho(p,d) \wedge \forall r \in data(d), reference(r) = name(t) : pairs(r) \\ \cup [name(p), defaultValue(p)]
\end{gathered}
} 
\\ \\
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge mandatory(p) \wedge instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered}
 \rho(p,d) \wedge \forall r \in row(owningClass(p)) : data(d)\\ \cup row(name(p), ["value", defaultValue(p)], \\ [name(p), value(primaryKey(r))]) 
\end{gathered}
} 
\end{cases}\\
p \in Properties, d \in Dataabses
\end{gather*}

@@

The $insertAssoc$ function is defined as:
\begin{gather*}
insertAssoc : Association \times Database \times (Row \times Row) \\
insertAssoc(a, d, f_m) = \begin{cases}
\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = name(owningClass(a)) \\ \wedge name(t_t) = reference(a) \\ \wedge \not \exists f \in foreignKeys(t_t) : name(f) = name(a) \\ startCardinality(a) \leq 1}{\begin{gathered}
\rho(a,d) \wedge \forall r \in f_m(owningClass(a), reference(a)) : \\ data(t_t) \cup row(name(t_s), [name(a), \\ value(primaryKey(first(r)))]) 
\end{gathered}
 }
 \\ \\
\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = name(owningClass(a)) \\ \wedge name(t_t) = reference(a) \\ \wedge \not \exists f \in foreignKeys(t_t) : name(f) = name(a) \\ startCardinality(a) > 1}{\begin{gathered}
\rho(a,d) \wedge \forall r \in f_m(owningClass(a), reference(a)),\\ t \in tables(d) : name(t) = name(a) : \\ data(t) \cup row(name(a), [name(t_s), \\ value(primaryKey(first(r)))] [name(t_t), \\ value(primaryKey(second(r)))])
\end{gathered}
 } 
 \end{cases}\\
 a \in Associations, d \in Databases
\end{gather*}

\begin{gather*}
\mathbf{newClass} : Class \times State \rightarrow State \\
\Psi(s, newClass(c)) = application(s) \circ addClass(c) \\
\Phi(s, newClass(c)) = database(s) \bullet insertClass(c) 
\end{gather*}
\begin{gather*}
addClass: Class \times Application \rightarrow Application \\ 
addClass(c, a) = \inference{\forall s \in classes(a) : name(s) \neq name(c)}{classes(a) \cup c} \\ \\
c \in Classes, a \in Applications
\end{gather*}
\paragraph{Add Class} The $insertClass$ operation changes the structure of the database schema only.
\begin{gather*}
insertClass : Class \times Database \rightarrow Database \\
insertClass(c, d) = \inference{ \forall t \in tables(d) : name(t) \neq name(\rho(c)) \\
	\forall a \in associations(c), \exists u \in  tables(d) : reference(a) = name(u) 
}{
\begin{gathered}
  \rho(c) \wedge \rho(singleProperties(c)) \wedge \rho(collectionProperties(c)) \\ \wedge \rho(associations(c))
\end{gathered}
}
\end{gather*}

\subsection{Software Decomposition}
\subsection{Software Alternation}

\subsubsection{Application Deconstruction}
\begin{gather*}
RemoveClass: Label \times Application \rightarrow Application  \\
RemoveClass(q, a) = \inference{\exists c \in classes(a) : name(c) = q \\
\wedge !\: instantiated(c) = \varnothing \wedge !\: associated(c)} {classes(a) \setminus c } \\
q \in Label, a \in Applications
\end{gather*}
\begin{gather*}
RemoveProperty: Label \times Label \times Application \rightarrow Application \\
RemoveProperty(l_c, l_p, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists p \in properties(c) : name(p) = l_p}{properties(c) \setminus p }\\
l_c, l_p \in Label, a \in Applications 
\end{gather*}
\begin{gather*}
RemoveAssociation : Label \times Labels \times Application \rightarrow Application \\ 
RemoveAssociation(l_c, l_a, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists s \in associations(c) : name(s) = l_a }{associations(c) \setminus s} \\
l_c, l_a \in Labels, a \in Applications
\end{gather*}

\subsubsection{Application Alternation}
\paragraph{Rename Class}
\begin{gather*}
\rho_{RenC} : Label \times Label \times Application \rightarrow Application \\ 
\rho_{RenC}(l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \forall e \in classes(a) : name(e) \neq l_n}{\begin{gathered}
c.name = l_n \wedge \forall f \in associations(e), \\ e \in classes(a), reference(f) = l_c : f.reference = l_n 
\end{gathered}}\\
l_o, l_n \in Label, a \in Applications
\end{gather*}

@@

\subsection{Database Transformation}

\subsubsection{Deconstruction of a Software}
\paragraph{Remove Class}
$$
\Phi_{RC}(c,d) = \inference{ \rho(c) \in tables(d) \\ \wedge \not \exists r \in data(d) : owningTable(r) = \rho(c) \\ \not \exists j \in tables(d), f \in foreignKeys(j) : reference(f) = \rho(c)  }{ tables(d) \setminus \rho(c)}
$$
%TODO remove by label

\paragraph{Remove Property}
$$
\Phi_{RP}(p, d) = \begin{cases}
 \inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge \exists c \in columns(t) : name(c) = name(p) \\ \wedge \not \exists r \in data(d) : owningTable(r) = name(owningClass(p)) }{columns(t) \setminus c
} \\ \\
  \inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge \not \exists c \in columns(t) : name(c) = name(p) \\ \wedge \exists u \in tables(d) : name(u) = name(p) \\ \wedge \not \exists r \in data(d) : owningTable(r) = name(owningClass(p))}{ tables(d) \setminus u }
 \end{cases}
$$

\paragraph{Remove Association}
$$
\Phi_{RA}(a, d) =  \begin{cases}
 \inference{\exists t \in tables(d) : name(t) = name(owningClass(a))
 \\ \wedge \exists f \in foreignKeys(t) : name(f) = name(a)
 \\ \wedge \forall r \in data(d) : owningTable(r) \neq t}{ foreignKeys(t) \setminus f }
 \\ \\
 \inference{\exists t \in tables(d) : name(t) = name(a)
 \\ \wedge \exists f \in foreignKeys(t) : name(f) = name(owningClass(a))
 \\ \wedge \forall r \in data(d) : owningTable(r) \neq t}{ tables(d) \setminus t}
 \end{cases}
$$

\subsubsection{Altering a Software}	
\paragraph{Copy Column}
$$\Phi_{copyC} : Label \times Label \times Label \times f_m \times Database \rightarrow Database $$

\subsection{Composing of Evolution Operations}
\paragraph{Copy Property}
All transformations described earlier does change only the structure of a application. These operation can be used in case there are no instances in the application. If instances are present in the application, the decomposition transformations could cause a data lost. Therefore we define  the transformation for copying a property from one class to another. The copy property is the basic way to copy data from instance of a class to instance of another class.

@@

>>>>

$\rho_p(p, d) = \begin{cases}
  \inference{cardinality(p) = 1 \wedge !\;mandatory(p)
 \\ \wedge \exists t : t = table(name(owningClass(p)), d) 
 \\ \wedge \not instanciated(owningClass(p))
 }{\begin{gathered}
	  column(name(p), \rho_t(type(p)), \varnothing )
\end{gathered}
 } \\ \\

  \inference{cardinality(p) = 1 \wedge mandatory(p)
 \\ \wedge \exists t : t = table(name(owningClass(p)), d)}{\begin{gathered}
 	 column(name(p), \rho_t(type(p)), NOTNULL ) \\ 
\end{gathered}} 
\\ \\

@@

 \inference{ cardinality(p) > 1 \wedge mandatory(p) \\ \wedge \exists t : t = table(name(owningClass(p)), d) }{\begin{gathered}  table(name(p), primaryKey("Id"), \\ column("value", \rho_t(type(p)), NOTNULL), \\ foreignKey(name(p), name(owningClass(p)), NOTNULL)
	 \end{gathered}}
 \end{cases}$

@@

\subsection{Software Creation}
There are defined three transformations which are able to create a software: $newClass$, $addProperty$, $addAssociation$. These three transformations can be used to create a software from a scratch. The first of them has no impact on stored data, whereas the rest of them can in some cases.
\begin{gather*}
\mathbf{addProperty} : Label \times Property \times State \rightarrow State \\
\Psi(s, addProperty(l, p)) = application(s) \circ addProp(l,p) \\
\Phi(s, addProperty(l, p)) = database(s) \bullet insertProp(l, p) \\
s \in Softwares, l \in Label, p \in Properties
\end{gather*}
Adding a property into an application structure has to verify a name collision only:
\begin{gather*}
addProp : Label \times Property \times Application \rightarrow Application \\
addProp(l, p, a) = \inference{\exists c \in classes(a) : name(c) = l \\ \wedge \forall q \in  properties(p) : name(q) \neq name(p)}{properties(c) \cup p } \\
l \in Label, p \in Properties, a \in Applications
\end{gather*}
Whereas the $insertProp$ on database level can change stored data too, because a mandatory property can be added into a class with already existing instances, then a default value has to be added to all existing instances.
\begin{gather*}
insertProp(p, d) = \begin{cases}
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge (!\;mandatory(p) \vee !\;instantiated(owningClass(p))}{ columns(t) \cup \rho(p,d)} 
\\ \\ 
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge mandatory(p) \wedge instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered} columns(t) \cup 
 \rho(p,d) \\ \wedge \forall r \in data(d), reference(r) = name(t) : pairs(r) \\ \cup [name(p), defaultValue(p)]
\end{gathered}
} 
\\ \\
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge mandatory(p) \wedge instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered}
tables(d) \cup  \rho(p,d) \wedge \forall r \in row(owningClass(p)) : data(d)\\ \cup row(name(p), ["value", defaultValue(p)], \\ [name(p), value(primaryKey(r))]) 
\end{gathered}
} 
\end{cases}\\
p \in Properties, d \in Dataabses
\end{gather*}

@@

The $insertAssoc$ function is defined as:
\begin{gather*}
insertAssoc : Association \times Database \times (Row \times Row) \\
insertAssoc(a, d, f_m) = \begin{cases}
\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = name(owningClass(a)) \\ \wedge name(t_t) = reference(a) \\ \wedge \not \exists f \in foreignKeys(t_t) : name(f) = name(a) \\ startCardinality(a) \leq 1}{\begin{gathered}
columns(t_s) \cup \rho(a,d) \\ \wedge \forall r \in f_m(owningClass(a), reference(a)) : \\ data(t_t) \cup row(name(t_s), [name(a), \\ value(primaryKey(first(r)))]) 
\end{gathered}
 }
 \\ \\
\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = name(owningClass(a)) \\ \wedge name(t_t) = reference(a) \\ \wedge \not \exists f \in foreignKeys(t_t) : name(f) = name(a) \\ startCardinality(a) > 1}{\begin{gathered} tables(d) \cup
\rho(a,d) \\ \wedge \forall r \in f_m(owningClass(a), reference(a)),\\ t \in tables(d) : name(t) = name(a) : \\ data(t) \cup row(name(a), [name(t_s), \\ value(primaryKey(first(r)))] [name(t_t), \\ value(primaryKey(second(r)))])
\end{gathered}
 } 
 \end{cases}\\
 a \in Associations, d \in Databases
\end{gather*}

\begin{gather*}
\mathbf{newClass} : Class \times State \rightarrow State \\
\Psi(s, newClass(c)) = application(s) \circ addClass(c) \\
\Phi(s, newClass(c)) = database(s) \bullet insertClass(c) 
\end{gather*}
\begin{gather*}
addClass: Class \times Application \rightarrow Application \\ 
addClass(c, a) = \inference{\forall s \in classes(a) : name(s) \neq name(c)}{classes(a) \cup c} \\ \\
c \in Classes, a \in Applications
\end{gather*}
\paragraph{Add Class} The $insertClass$ operation changes the structure of the database schema only, because the $insertAssoc$ transformation needs a mapping function to work properly, which is not part of the $newClass$ definition. The situation reflect the modeling praxis when the structure of a class is defined and then it is associations are created.
\begin{gather*}
insertClass : Class \times Database \rightarrow Database \\
insertClass(c, d) = \inference{ \forall t \in tables(d) : name(t) \neq name(\rho(c)) \\
	%\forall a \in associations(c), \exists u \in  tables(d) : reference(a) = name(u) \\ 
	assocaiations(c) = \varnothing
}{
\begin{gathered}
  tables(d) \cup \rho(c) \\ \wedge \rho(singleProperties(c)) \wedge \rho(collectionProperties(c)) \\ 
\end{gathered}
}
\end{gather*}

\subsection{Software Decomposition}
\begin{gather*}
RemoveClass: Label \times Software \rightarrow Software  \\
\Psi RemoveClass(l, s) = applications(s) \circ remClass(l) \\
\Phi RemoveClass(l, s) = database(s) \bullet dropClass(l) \\
\\ l \in Label, s \in Software 
\\ \\
remClass(l, a) = \inference{\exists c \in classes(a) : name(c) = l \\
\wedge !\: instantiated(c) = \varnothing \wedge !\: associated(c)} {classes(a) \setminus c } \\
\end{gather*}
\begin{gather*}
dropClass(l,d) = \inference{ \exists t \in tables(d) : name(t) = l  \\ \wedge \not \exists r \in data(d) : owningTable(r) = l \\  \not \exists j \in tables(d) \wedge f \in foreignKeys(j) : reference(f) = l }{ tables(d) \setminus \rho(c)}
\end{gather*}

\begin{gather*}
RemoveProperty: Label \times Label \times Software \rightarrow Software \\
\Psi RemoveProperty(l, l_p, s) =  applications(s) \circ  remProp(l, l_p) \\
\Phi RemoveProperty(l, l_p, s) =  database(s) \bullet dropProp(l, l_p) \\
RemoveProperty(l_c, l_p, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists p \in properties(c) : name(p) = l_p}{properties(c) \setminus p }\\
l_c, l_p \in Label, a \in Applications 
\end{gather*}
\begin{gather*}
dropProp(p, d) = \begin{cases}
 \inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge \exists c \in columns(t) : name(c) = name(p) \\ \wedge \not \exists r \in data(d) : owningTable(r) = name(owningClass(p)) }{columns(t) \setminus c
} \\ \\
  \inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge \not \exists c \in columns(t) : name(c) = name(p) \\ \wedge \exists u \in tables(d) : name(u) = name(p) \\ \wedge \not \exists r \in data(d) : owningTable(r) = name(owningClass(p))}{ tables(d) \setminus u }
 \end{cases}
\end{gather*}

\begin{gather*}
RemoveAssociation : Label \times Labels \times Software \rightarrow Software \\ 
\Psi RemoveAssociation(l, l_a, s) = application(s) \circ remAss(l, l_a) \\
\Phi RemoveAssociation(l, l_a, s) = database(s) \bullet dropAssoc(l, l_a) \\
RemoveAssociation(l_c, l_a, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists f \in associations(c) : name(f) = l_a }{associations(c) \setminus f} \\
l_c, l_a \in Labels, a \in Applications
\end{gather*}
\begin{gather*}
dropAssoc(l, l_a, d) =  \begin{cases}
 \inference{\exists t \in tables(d) : name(t) = l
 \\ \wedge \exists f \in foreignKeys(t) : name(f) = l_a
 \\ \wedge \forall r \in data(d) : owningTable(r) \neq t}{ foreignKeys(t) \setminus f }
 \\ \\
 \inference{\exists t \in tables(d) : name(t) = l_a
 \\ \wedge \exists f \in foreignKeys(t) : name(f) = l
 \\ \wedge \forall r \in data(d) : reference(r) \neq t}{ tables(d) \setminus t}
 \end{cases}
\end{gather*}

\subsection{Software Alternation}
\paragraph{Rename Class}
\begin{gather*}
\rho_{RenC} : Label \times Label \times Application \rightarrow Application \\ 
\rho_{RenC}(l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \forall e \in classes(a) : name(e) \neq l_n}{\begin{gathered}
c.name = l_n \wedge \forall f \in associations(e), \\ e \in classes(a), reference(f) = l_c : f.reference = l_n 
\end{gathered}}\\
l_o, l_n \in Label, a \in Applications
\end{gather*}

@@

\subsubsection{Altering a Software}	
\paragraph{Copy Property}
All transformations described earlier does change only the structure of a application. These operation can be used in case there are no instances in the application. If instances are present in the application, the decomposition transformations could cause a data lost. Therefore we define  the transformation for copying a property from one class to another. The copy property is the basic way to copy data from instance of a class to instance of another class.

@@

\paragraph{Copy Column}
$$\Phi_{copyC} : Label \times Label \times Label \times f_m \times Database \rightarrow Database $$

\subsection{Composing of Evolution Operations}
